---
title: "Plots for read stats"
output:
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: no
    code_folding: hide
params:
    map_tab: x
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)
```

```{r include=FALSE}
library(rmarkdown)
library(ggplot2)
library(dplyr)
library(tidyr)
library(stringr)
library(scales)
library(zoo)
library(httpgd)
```

# Load mapping table
```{r}
map_tab <- read.csv(params$map_tab, sep = '\t', header = TRUE)

# map_tab <- read.csv("/home/ubuntu/aav_integration/v1.0.0/data/interim/shortRead/240112_VH00163_77_AAF5V37M5__mapped_table.tsv", sep = '\t', header = TRUE)
map_tab$Sample <- sub("_a1", '', map_tab$Sample)

map_tab <- map_tab %>% 
    mutate(sorting_key = as.numeric(str_extract(Sample, "(?<=_S)\\d+$")),
            Sample = factor(Sample, levels = Sample[order(sorting_key)]))

data <- map_tab %>%
    mutate(unmapped_reads = tot_input - mixedGenome) %>%
    rename(mapped_reads = mixedGenome)

# convert to reads per million
data_rpm <- data %>% 
    select(2, 4, 11, 13:21,23) %>%
    mutate(across(-matches("Sample"), ~ (. / tot_input) * 10^6))
```

# Fig 1: Total read numbers for each sample 
```{r}
ggplot(data, aes(x = Sample, y = tot_input)) + 
    geom_bar(stat = "identity") +
    theme_minimal() +
    labs(x = "Sample", y = "Total reads", title = "# of Total reads") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, size = 20)) +
    scale_y_continuous(labels = scales::scientific)

```

# Fig 2: % mapped Fraction 
- raw reads  
```{r}

# Convert to long format
data_long <- data %>%
    select(Sample, mapped_reads, unmapped_reads) %>%
    pivot_longer(cols = c(mapped_reads, unmapped_reads),
                names_to = "type",
                values_to = "reads")

data_long$type <- factor(data_long$type, levels = c("unmapped_reads", "mapped_reads"))

ggplot(data_long, aes(x = Sample, y = reads, fill = type)) +
    geom_bar(stat = "identity", position = "stack") +
    # scale_y_continuous(labels = scales::percent) +
    labs(x = "Sample Names", y = "Reads", title = "Mapped and Unmapped Reads") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_manual(values = c("mapped_reads" = "#026d02", "unmapped_reads" = "#d70707")) +
    scale_y_continuous(labels = scales::scientific)
```

- scaled to 100%  
```{r}
ggplot(data_long, aes(x = Sample, y = reads, fill = type)) +
    geom_bar(stat = "identity", position = "fill") +
    # scale_y_continuous(labels = scales::percent) +
    labs(x = "Sample Names", y = "Percentage of reads", title = "Mapped and Unmapped Reads") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_manual(values = c("mapped_reads" = "#026d02", "unmapped_reads" = "#d70707")) +
    scale_y_continuous(labels = scales::percent)


```


# Fig 3. vector reads
- raw reads  
```{r}
ggplot(data, aes(x = Sample, y = vectorGenome)) + 
    geom_bar(stat = "identity") +
    theme_minimal() +
    labs(x = "Sample", y = "Reads", title = "# of reads mapped to vector") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, size = 20)) +
    scale_y_continuous(labels = scales::scientific)
```

- normalized to total reads (reads per million)  
```{r}
# reads per million
ggplot(data_rpm, aes(x = Sample, y = vectorGenome)) + 
    geom_bar(stat = "identity") +
    theme_minimal() +
    labs(x = "Sample", y = "Reads per million", title = "Reads mapped to vector (RPM)") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, size = 20)) +
    scale_y_continuous(labels = scales::scientific)
```

# Fig 4: % reads after filters:  
- 1, remove PCR duplicates  
- 2, remove non properly paired reads on host genome 
- 3, remove non uniquely mapped reads (only for v1.1.1 pipeline)  

```{r}
df_long <- data %>% select(Sample, mapped_reads, starts_with("mixedGenome_")) %>% 
    mutate(across(starts_with("mixedGenome_"), ~ . / mapped_reads * 100)) %>%
    pivot_longer(cols = starts_with("mixedGenome_"), names_to = "category", values_to = "frac_reads")

df_long$category <- gsub("mixedGenome_", "", df_long$category)

df_long$category <- factor(df_long$category, levels = c("rmdup", "rmdup_paired"))

ggplot(df_long, aes(x = Sample, y = frac_reads, fill = category)) +
    geom_bar(stat = "identity", position = position_dodge()) +
    scale_fill_brewer(palette = "Pastel1") +
    labs(x = "Sample", y = "% reads remained", title = "Reads remained after filters") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 45, hjust = 1))

```


# Fig 5: subtypes of chimeric reads
- raw reads  
```{r}
df_long <- data %>% select(Sample, vector_vector, host_host, vector_host) %>%
    pivot_longer(cols = !"Sample", names_to = "subtype", values_to = "Reads")

ggplot(df_long, aes(x = Sample, y = Reads, fill = subtype)) +
    geom_bar(stat = "identity", position = "stack") +
    scale_fill_brewer(palette = "Pastel1") + # Optional: Use a color palette
    labs(x = "Sample", y = "Reads", title = "Subtype of chimeric reads") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 45, hjust = 1))

```

- scaled to 100%  
```{r}
df_long <- data %>% select(Sample, vector_vector, host_host, vector_host) %>%
    pivot_longer(cols = !"Sample", names_to = "subtype", values_to = "Reads")

ggplot(df_long, aes(x = Sample, y = Reads, fill = subtype)) +
    geom_bar(stat = "identity", position = "fill") +
    scale_fill_brewer(palette = "Pastel1") + # Optional: Use a color palette
    labs(x = "Sample", y = "Percentage of read", title = "Subtype of chimeric reads") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 45, hjust = 1))

```
